\mytitlestring

### 2013-05-31 trunk/fqas2013/paper.tex
Laatste kleine aanpassingen voor indienen.

\newcommand{\mytitlestring}{Bipolar Querying of Valid-time Intervals Subject to Uncertainty}

@@

\section{Introduction}
Generally, database systems model (parts of) reality. For this, their databases contain data representing properties of real-world objects or concepts~\cite{Pons2012ijcis},~\cite{Billiet2012ipmu},~\cite{Pons2012ipmu}. Some essential properties of real-world objects or concepts are time-related. Thus, databases often contain data representing temporal values~\cite{Bohlen1998lncs}, which are basically indications of time and describe such properties~\cite{Pons2012ijcis},~\cite{Jensen1999ieeetkde},~\cite{Galindo2001}. These temporal values are usually either time intervals~\cite{Bohlen1998lncs} or instants~\cite{Bohlen1998lncs}, which may informally be seen as infinitesimally short `periods' or `points' in time. Based on their interpretation and purpose, temporal values can be classified into several categories, but the presented work will only consider valid-time indications, which indicate when corresponding data is a valid or true representation of the reality modelled by its database~\cite{Bohlen1998lncs},~\cite{Pons2012ijcis},~\cite{Billiet2012ipmu},~\cite{Pons2012ipmu}.

A lot of database data are produced by humans, but human-made data are prone to imperfections, as some of these data may be vague, imprecise, contradictory, incomplete or uncertain~\cite{Billiet2012ipmu},~\cite{Pons2012ipmu},~\cite{Medina1994is},~\cite{Bosc2010ijufkbs}. Of course, data representing temporal values may contain such imperfections too~\cite{Billiet2012ipmu},~\cite{Pons2012ipmu},~\cite{Pons2012ijcis},~\cite{Dyreson1998acm}. The work presented in this paper will consider databases containing data representing valid-time intervals subject to uncertainty and will assume all non-temporal data in these databases to contain no imperfections.

One of the most important purposes of a database is to allow its data to be queried, to allow the information or knowledge represented by this data to be retrieved. A user may query a database in a `regular' way: the user describes the data which he or she finds desired or satisfactory and thus wants to retrieve, by perfectly describing the allowed values of these data. A user may also query a database in a `fuzzy' way: the user describes the data which he or she finds desired or satisfactory by imperfectly describing the allowed values of these data~\cite{Zadrozny2008hrfipd}. These imperfect descriptions may contain vagueness or imprecision, often through the use of linguistic terms ~\cite{Devos1998jql},~\cite{Kacprzyk2001is}. A user may also query a database in a `bipolar' way. Generally, two main approaches to this exist. One is for the user to describe the data which he or she finds acceptable and to describe the data among this acceptable data, which he or she finds really desired, both by describing the allowed values of these data~\cite{Dubois2002lnai}. The other is for the user to independently describe both the data which he or she finds desired or satisfactory and the data which he or she finds undesired or unsatisfactory, both by describing the allowed values of these data~\cite{DeTre2010ieeetfs},~\cite{Matthe2011ijis}. The descriptions used in bipolar querying may contain imperfections. The presented work will only consider the latter approach to bipolar querying and will allow a simple form of imprecision in non-temporal elementary query conditions.

Compared to non-temporal user preferences, temporal user preferences usually have an uncommon nature and interpretation and thus expressing them relies on uncommon mechanics in querying: users usually prefer using specific temporal operators to express temporal preferences. Hence, several proposals have considered specific sets of temporal operators~\cite{Galindo2001},~\cite{Pons2012ijcis},~\cite{Schockaert2008ieeetfs}, often based on the possible temporal relationships between two time indications~\cite{Galindo2001},~\cite{Pons2012ijcis},~\cite{Schockaert2008ieeetfs}. Such temporal relationships define semantically meaningful relationships with a temporal nature, between two time indications. In~\cite{Allen1983cacm}, a collection of temporal relationships between two time intervals (and as a special case instants) is introduced and this collection is considered groundbreaking. Of course, to query temporal data in a fuzzy way, fuzzy variants of such temporal operators are necessary and several proposals have thus introduced such variants~\cite{Galindo2001},~\cite{Pons2012ipmu}, often based on fuzzy variants of such temporal relationships~\cite{Schockaert2008ieeetfs},~\cite{Pons2013ijufkbs}.

Techniques for the regular or fuzzy querying of valid-time databases containing valid-time data subject to uncertainty are considered by several existing proposals~\cite{Pons2012ijcis},~\cite{Pons2012ipmu},~\cite{Pons2013ijufkbs}. However, to the knowledge of the authors, few proposals have considered the bipolar querying of valid-time databases~\cite{Billiet2011fqas} and even fewer the bipolar querying of valid-time databases containing temporal data subject to uncertainty. Thus, this paper presents a novel technique to query a valid-time relation containing valid-time data subject to uncertainty in a bipolar way, which allows the user to specify a single bipolar temporal query condition. This paper is structured as follows: in section \ref{sec:preliminaries}, some preliminary concepts and techniques are described, in section \ref{sec:proposal}, the novel technique which is the main contribution of the work presented in this paper, is presented and in section \ref{sec:conclusions}, the conclusions of this paper and some directions for future research are given.

@@

The work presented in this paper concerns time indications subject to uncertainty. This uncertainty is always assumed to be caused by a (partial) lack of knowledge: the exact, intended time indication is not known, eventhough there is only one time indication intended and as such no variability. Confidence about exactly which time indication is the intended one in the context of such uncertainty is modelled using possibility theory~\cite{Dubois1988},~\cite{Pons2013ijufkbs}. In the presented work, `possibility' and `necessity' are always interpreted as measures of plausibility, respectively necessity, given all available knowledge. Time intervals not subject to any imperfection are called \emph{crisp time intervals} (CTI) in this paper.

\subsection{Valid-time Relations}
The presented work uses the relational database model~\cite{Pons2012ijcis},~\cite{Galindo2001}. Here, a \emph{relation} models a collection of similar real-world concepts or objects. To achieve this, such a relation has a set of $n$ \emph{attributes}, of which each describes a different property shared by all of these concepts or objects, and a set of $n$-tuples, of which each represents one of these real-world objects or concepts and contains $n$ values, each representing the capacity of the object or concept corresponding to the tuple, of the property described by the attribute corresponding to the value~\cite{Billiet2012ipmu},~\cite{Pons2012ipmu}.

A \emph{valid-time relation} (VTR) always has \emph{valid-time attributes}. These are attributes describing a single valid time~\cite{Bohlen1998lncs} for the objects or concepts represented by the VTR's tuples. A VTR may contain different tuples corresponding to the same real-world concept or object. The non-valid-time attribute values of such a tuple represent the capacities of the properties described by their corresponding attributes which were, are or will be true or valid for the object or concept corresponding to the tuple during the period in time indicated by the valid-time indication represented by the tuple's valid-time attribute values. Thus, such a tuple represents the `version' of the object or concept corresponding to this tuple which was, is or will be real or valid or the `state' this object or concept was, is or will be in, during the period in time indicated by the valid-time indication represented by the tuple's valid-time attribute values. In the presented work, such valid-time indications will always be time intervals~\cite{Bohlen1998lncs} and will always be referred to as \emph{valid-time intervals} (VTI).

\subsection{Uncertainty in Valid-time Intervals}
The presented work allows VTI to be subject to uncertainty, by allowing them to be \emph{ill-known valid-time intervals} (IKVTI). The concept of IKVTI is based on the concepts of \emph{possibilistic variables} (PV) and \emph{ill-known intervals} (IKI)~\cite{Dubois1988cma},~\cite{Pons2013ijufkbs},~\cite{Billiet2012ipmu}:

@@

Now consider a set $U$ containing single values (and not collections of values). When a PV $X_{v}$ is defined on such a set $U$, the unique value $X_{v}$ takes, which is (partially) unknown, will be a single value in $U$ and is called an \emph{ill-known value} (IKV) in $U$ in the presented work~\cite{Dubois1988cma},~\cite{Pons2013ijufkbs},~\cite{Billiet2012ipmu}. In this paper, IKV will be denoted using lower-case letters. The work presented in this paper uses a specific kind of IKI, defined as follows, although other definitions exist~\cite{Pons2012ipmu},~\cite{Billiet2012ipmu}:

@@

Specifically concerning valid time, an IKI in a time domain represented by the domains of a VTR's valid-time attributes is called an \emph{ill-known valid-time interval} (IKVTI) in the presented work. The work presented in this paper requires the possibility distributions defining an IKVTI's IKV to be convex~\cite{Pons2013ijufkbs}. In this paper, an IKVTI with boundary IKV $s$ and $e$ will be noted $\left[s, e\right]$.

@@

\begin{definition}
Given an ordered set $U$, an \emph{ill-known constraint} (IKC) $C$ on $U$ is specified by means of a binary relation $R \subseteq U^{2}$ and a fixed IKV $v$ in $U$, i.e.:
\begin{align}
C \triangleq (R,v) \nonumber
\end{align}
Any set $A \subseteq U$ now satisfies IKC $C = (R,v)$ if and only if:
\begin{align}
\forall a \in A : (a,v) \in R \nonumber
\end{align}
\end{definition}

The satisfaction of an IKC $C$ by a set $A$ will be noted $C(A)$ in this paper. Consider an ordered set $U$, an IKC $C \triangleq (R,v)$ on $U$ and a set $A \subseteq U$. Due to the uncertainty inherent to $v$, it is uncertain whether $A$ satisfies $C$ or not. The degree of possibility $\Pos(C(A))$ that $A$ satisfies $C$ and the degree of necessity $\Nec(C(A))$ that $A$ satisfies $C$, can be calculated as follows~\cite{Pons2013ijufkbs}:

\vspace{-10pt}
\begin{align}
\Pos(C(A)) & = \min_{a \in A}\left(\sup_{(a,w) \in R}\pi_{V}(w)\right) \label{ill-known-pos}\\
\Nec(C(A)) & = \min_{a \in A}\left(\inf_{(a,w) \notin R} 1-\pi_{V}(w)\right) \label{ill-known-nec}
\end{align}

Given an ordered set $U$, degrees of possibility and necessity that a set $A \subseteq U$ satisfies a boolean combination of IKC on $U$ can be found by using the possibilistic extensions of boolean operators `and' ($\wedge$), `or' ($\vee$) and `not' ($\neg$), as described in~\cite{Pons2013ijufkbs},~\cite{Pons2012ipmu},~\cite{Billiet2012ipmu}.

@@

\begin{table}[ht]
\caption{The translations of Allen relationships to the IKC framework.}
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Allen Relationship} & \textbf{Constraints} & \textbf{Combination} \\
\hline
\hline
I before J & $C_1\stackrel{\triangle}{=} \left(<,s\right)$ & $C_1(I)$ \\
\hline
\multirow{2}{*}
{I equal J} & $C_1\stackrel{\triangle}{=} \left(\geq,s\right)$, $C_2\stackrel{\triangle}{=} \left(\neq,s\right)$ & $C_1(I)\wedge$ $\neg C_2(I)\wedge$\\
& $C_3\stackrel{\triangle}{=} \left(\leq,e\right)$, $C_4\stackrel{\triangle}{=} \left(\neq,e\right)$ & $C_3(I)\wedge$ $\neg C_4(I)$\\
\hline
I meets J & $C_1\stackrel{\triangle}{=} \left(\leq,s\right)$ $C_2\stackrel{\triangle}{=} \left(\neq,s\right)$ & $C_1(I)\wedge$ $\neg C_2(I)$\\
\hline
I overlaps J & $C_1\stackrel{\triangle}{=} \left(<,e\right)$, $C_2\stackrel{\triangle}{=} \left(\leq,s\right)$, $C_3\stackrel{\triangle}{=} \left(\geq,s\right)$ & $C_1(I)\wedge$ $\neg C_2(I)\wedge$ $\neg C_3(I)$\\
\hline
\multirow{2}{*}
{I during J} & $C_1\stackrel{\triangle}{=} \left(>,s\right)$, $C_2\stackrel{\triangle}{=} \left(\leq,e\right)$ & $\big(C_1(I)\wedge$ $ C_2(I)\big)\vee$\\
 & $C_3\stackrel{\triangle}{=} \left(\geq,s\right)$, $C_4\stackrel{\triangle}{=} \left(<,e\right)$ & $\big(C_3(I)\wedge$ $C_4(I)\big)$\\
\hline
I starts J & $C_1\stackrel{\triangle}{=} \left(\geq,s\right)$, $C_2\stackrel{\triangle}{=} \left(\neq,s\right)$ & $C_1(I)\wedge$, $\neg C_2(I)$\\
\hline
I finishes J & $C_1\stackrel{\triangle}{=} \left(\leq,e\right)$, $C_2\stackrel{\triangle}{=} \left(\neq,e\right)$ & $C_1(I)\wedge$ $\neg C_2(I)$\\
\hline
\end{tabular}
\label{tab:allen-relations}
\end{table}

\subsection{Bipolar Querying}
As mentioned before, humans may express their query preferences using both positive and negative query conditions~\cite{Matthe2011ijis},~\cite{DeTre2010ieeetfs}. If the semantics of these conditions are non-symmetric, meaning that the positive preferences can not be derived from the negative or vice versa, the bipolarity in this query is called \emph{heterogenous}~\cite{Matthe2011ijis}. The presented work will concern only such heterogenous query bipolarity.

Bipolarity in a query can either be specified between or inside the elementary query conditions. In~\cite{Matthe2011ijis}, it is shown that combining both approaches makes no sense and that the approach where bipolarity is specified inside elementary query conditions, using intuitionistic fuzzy sets~\cite{Atanassov1986fss}, is a more intuitive one. In the presented work, only the latter approach is used. In this approach, elementary query conditions express both what is accepted and what is not accepted by the query, at once, and are called \emph{bipolar query conditions} (BQC)~\cite{Matthe2011ijis}.

Consider a relation attribute $A$. Let $dom_{A}$ be the domain of $A$'s data type, let $\mu_{c_{A}}$ and $\nu_{c_{A}}$ be membership functions from $dom_{A}$ to the unit interval $\left[0, 1\right]$, where $\mu_{c_{A}}(x)$ represents to what extent $x \in dom_{A}$ is satisfactory and $\nu_{c_{A}}(x)$ to what extent $x$ is unsatisfactory to a user, then a BQC $c_{A}$ expressing this user's preferences concerning $A$ can be modelled by an Intuitionistic Fuzzy Set (IFS)~\cite{Atanassov1986fss} as~\cite{Matthe2011ijis},~\cite{DeTre2010ieeetfs}:
\vspace{-5pt}
\begin{equation}
c_{A} = \{(v, \mu_{c_{A}}(v), \nu_{c_{A}}(v)) : v \in dom_{A}\}
\end{equation}

Note that to allow overspecification of the user's preferences, the IFS's consistency condition can be relaxed, which means that there may exist values $v \in dom_{A}$ for which $\mu_{c_{A}}(v) + \nu_{c_{A}}(v) > 1$~\cite{Matthe2011ijis},~\cite{DeTre2010ieeetfs}.

@@

In general, the evaluation of a BQC $c_{A}$ on relation attribute $A$ for a relation tuple $R$ will result in a BSD $(s_{c_A}^R, d_{c_A}^R)$, which is calculated as follows. Let $R[A]$ denote the value of tuple $R$ for attribute $A$, then~\cite{Matthe2011ijis}:
\begin{equation}\label{eq:bsdeval}
(s_{c_A}^R, d_{c_A}^R)=(\mu_{c_A}(R[A]), \nu_{c_A}(R[A]))
\end{equation}

@@

\subsection{Valid-time Relations Subject to Uncertainty}
The presented proposal will concern VTR where the VTI are IKVTI. Generally, such a VTR $R$ can be seen as constructed in the following way. Let $R$ have $n$ non-temporal attributes $A_{i}, 1 \leq i \leq n, i \in \mathbb{N}$ and two valid-time attributes $VST$ and $VET$. Every tuple $T$ of $R$ represents an object or concept version or state which is valid during the time period indicated by the tuple's IKVTI $I_T$. This IKVTI $I_T$ is now defined by two IKV, which respectively describe the starting and ending instants of $I_T$ and are represented by the tuple's VST, respectively VET values. The interpretation is that the version or state corresponding to a tuple was, is or will be valid during a period of time, but exactly which period this is intended to be, is unknown. Confidence about exactly which period is intended, is modelled by the tuple's IKVTI~\cite{Pons2012ijcis},~\cite{Pons2012ipmu}.

\vspace{-10pt}
\begin{table}[ht]
\caption{The example relation used in this paper.}
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{ID} & \textbf{Author} & \textbf{VST} & \textbf{VET} \\
\hline
\hline
1 & Alo\"{i}sius & $\left[4/4/1208, 6/4/1208, 16/4/1208\right]$ & $\left[10/12/1208, 1/1/1209, 26/1/1209\right]$ \\
\hline
2 & Theofilus & $\left[2/4/1209, 12/4/1209, 22/4/1209\right]$ & $\left[21/12/1209, 1/1/1210, 21/1/1210\right]$ \\
\hline
3 & Gerardus & $\left[14/1/1209, 15/1/1209, 16/1/1209\right]$ & $\left[21/12/1209, 15/1/1210, 25/1/1210\right]$ \\
\hline
4 & Euforius & $\left[21/12/1210, 1/1/1211, 11/1/1211\right]$ & $\left[21/12/1211, 1/1/1212, 11/1/1212\right]$ \\
\hline
5 & Ambrosius & $\left[11/12/1213, 21/12/1213, 15/1/1214\right]$ & $\left[9/10/1216, 10/10/1216, 15/10/1216\right]$ \\
\hline
6 & Alo\"{i}sius & $\left[21/12/1213, 1/1/1214, 11/1/1214\right]$ & $\left[9/6/1217, 9/6/1217, 12/6/1217\right]$ \\
\hline
7 & Gerardus & $\left[29/12/1214, 1/1/1215, 8/1/1215\right]$ & $\left[9/6/1217, 10/6/1217, 12/6/1217\right]$ \\
\hline
\end{tabular}
\label{tab:ex-relation}
\end{table}
\vspace{-10pt}

@@

\vspace{-10pt}
\begin{align}
\pi(x) =
\begin{cases}
\frac{x - d_1}{d_2 - d_1}, & \text{if } d_1 \leq x < d_2 \\
\frac{d_3 - x}{d_3 - d_2}, & \text{if } d_2 \leq x \leq d_3 \\
0, & \text{else}
\end{cases}
\end{align}

>>>>

\newcommand{\mytitlestring}{Bipolar Querying of Valid-Time Intervals Subject to Uncertainty}

@@

\section{Introduction}
Generally, database systems model (parts of) reality. For this, their databases contain data representing properties of real-world objects or concepts. Some essential properties of real-world objects or concepts are time-related. Thus, databases often contain data representing temporal values~\cite{Bohlen1998lncs}, which are basically indications of time and describe such properties. These temporal values are usually either time intervals~\cite{Bohlen1998lncs} or instants~\cite{Bohlen1998lncs}, which may informally be seen as infinitesimally short `periods' or `points' in time. Based on their interpretation and purpose, temporal values can be classified into several categories, but the presented work will only consider valid-time indications, which indicate when corresponding data is a valid or true representation of the reality modelled by its database~\cite{Bohlen1998lncs,Pons2012ijcis,Billiet2012ipmu,Pons2012ipmu}.

A lot of database data are produced by humans, but human-made data are prone to imperfections, as some of these data may be vague or imprecise~\cite{Medina1994is}, contradictory, incomplete or uncertain~\cite{Billiet2012ipmu},~\cite{Pons2012ipmu},~\cite{Bosc2010ijufkbs}. Of course, data representing temporal values may contain such imperfections too~\cite{Pons2012ijcis,Billiet2012ipmu,Pons2012ipmu},~\cite{Dyreson1998acm}. The work presented in this paper will consider databases containing data representing valid-time intervals subject to uncertainty and will assume all non-temporal data in these databases to contain no imperfections.

One of the most important purposes of a database is to allow its data to be queried, to allow the information or knowledge represented by this data to be retrieved. A user may query a database in a `regular' way: the user describes the data which he or she finds desired or satisfactory and thus wants to retrieve, by perfectly describing the allowed values of these data. A user may also query a database in a `fuzzy' way: the user describes the data which he or she finds desired or satisfactory by imperfectly describing the allowed values of these data~\cite{Zadrozny2008hrfipd}. These imperfect descriptions may contain vagueness or imprecision, often through the use of linguistic terms~\cite{Devos1998jql},~\cite{Kacprzyk2001is}. A user may also query a database in a `bipolar' way. Generally, two main approaches to this exist. One is for the user to describe the data which he or she finds acceptable and to describe the data among this acceptable data, which he or she finds really desired, both by describing the allowed values of these data~\cite{Dubois2002lnai}. The other is for the user to independently describe both the data which he or she finds desired or satisfactory and the data which he or she finds undesired or unsatisfactory, both by describing the allowed values of these data~\cite{DeTre2010ieeetfs},~\cite{Matthe2011ijis}. The descriptions used in bipolar querying may contain imperfections. The presented work will only consider the latter approach to bipolar querying and will allow a simple form of imprecision in non-temporal elementary query conditions.

Compared to non-temporal user preferences, temporal user preferences usually have an uncommon nature and interpretation and thus expressing them relies on uncommon mechanics in querying: users usually prefer using specific temporal operators to express temporal preferences. Hence, several proposals have considered specific sets of temporal operators, often based on the possible temporal relationships between two time indications~\cite{Pons2012ijcis},~\cite{Galindo2001},~\cite{Schockaert2008ieeetfs}. Such temporal relationships define semantically meaningful relationships with a temporal nature, between two time indications. In~\cite{Allen1983cacm}, a collection of temporal relationships between two time intervals (and as a special case instants) is introduced and this collection is considered groundbreaking. Of course, to query temporal data in a fuzzy way, fuzzy variants of such temporal operators are necessary and several proposals have thus introduced such variants~\cite{Pons2012ipmu},~\cite{Galindo2001}, often based on fuzzy variants of such temporal relationships~\cite{Schockaert2008ieeetfs},~\cite{Pons2013ijufkbs}.

Techniques for the regular or fuzzy querying of valid-time databases containing valid-time data subject to uncertainty are considered by several existing proposals~\cite{Pons2012ijcis},~\cite{Pons2012ipmu},~\cite{Pons2013ijufkbs}. However, to the knowledge of the authors, only one proposal has considered the bipolar querying of valid-time databases (in~\cite{Billiet2011fqas}, a technique is proposed to query a valid-time database containing temporal data subject to imprecision in a bipolar way) and none have considered the bipolar querying of valid-time databases containing temporal data subject to uncertainty. Thus, this paper presents a novel technique to query a valid-time relation containing valid-time data subject to uncertainty in a bipolar way, which allows the user to specify a single bipolar temporal query condition. This paper is structured as follows: in section \ref{sec:preliminaries}, some preliminary concepts and techniques are described, in section \ref{sec:proposal}, the novel technique which is the main contribution of the work presented in this paper, is explained and in section \ref{sec:conclusions}, the conclusions of this paper and some directions for future research are given.

@@

The work presented in this paper concerns time indications subject to uncertainty. This uncertainty is always assumed to be caused by a (partial) lack of knowledge: the exact, intended time indication is not known, eventhough there is only one time indication intended and as such no variability. Confidence about exactly which time indication is the intended one in the context of such uncertainty is modelled using possibility theory~\cite{Pons2013ijufkbs},~\cite{Dubois1988}. In the presented work, `possibility' and `necessity' are always interpreted as measures of plausibility, respectively necessity, given all available knowledge. Time intervals not subject to any imperfection are called \emph{crisp time intervals} (CTI) in this paper.

\subsection{Valid-Time Relations}
A \emph{valid-time relation} (VTR) always has \emph{valid-time attributes}. These are attributes describing a single valid time~\cite{Bohlen1998lncs} for the objects or concepts represented by the VTR's tuples. A VTR may contain different tuples corresponding to the same real-world concept or object. The non-valid-time attribute values of such a tuple represent the capacities of the properties described by their corresponding attributes which were, are or will be true or valid for the object or concept corresponding to the tuple during the period in time indicated by the valid-time indication represented by the tuple's valid-time attribute values. Thus, such a tuple represents the `version' of the object or concept corresponding to this tuple which was, is or will be real or valid or the `state' this object or concept was, is or will be in, during the period in time indicated by the valid-time indication represented by the tuple's valid-time attribute values. In the presented work, such valid-time indications will always be time intervals~\cite{Bohlen1998lncs} and will always be referred to as \emph{valid-time intervals} (VTI).

\subsection{Uncertainty in Valid-Time Intervals}
The presented work allows VTI to be subject to uncertainty, by allowing them to be \emph{ill-known valid-time intervals} (IKVTI). The concept of IKVTI is based on the concepts of \emph{possibilistic variables} (PV) and \emph{ill-known intervals} (IKI)~\cite{Billiet2012ipmu},~\cite{Pons2013ijufkbs},~\cite{Dubois1988cma}.

@@

Now consider a set $U$ containing single values (and not collections of values). When a PV $X_{v}$ is defined on such a set $U$, the unique value $X_{v}$ takes, which is (partially) unknown, will be a single value in $U$ and is called an \emph{ill-known value} (IKV) in $U$~\cite{Billiet2012ipmu},~\cite{Pons2013ijufkbs},~\cite{Dubois1988cma}. In this paper, IKV will be denoted using lower-case letters. The work presented in this paper uses a specific kind of IKI, defined as follows, although other definitions exist~\cite{Billiet2012ipmu},~\cite{Pons2012ipmu}.

@@

Specifically concerning valid time, an IKI in a time domain represented by the domains of a VTR's valid-time attributes is called an \emph{ill-known valid-time interval} (IKVTI). The work presented in this paper requires the possibility distributions defining an IKVTI's IKV to be convex~\cite{Pons2013ijufkbs}. In this paper, an IKVTI with boundary IKV $s$ and $e$ will be noted $\left[s, e\right]$.

@@

\begin{definition}
Given an ordered set $U$, an \emph{ill-known constraint} (IKC) $C = (R,v)$ on $U$ is specified by means of a binary relation $R \subseteq U^{2}$ and a fixed IKV $v$ in $U$. Any set $A \subseteq U$ now satisfies IKC $C = (R,v)$ if and only if:
\begin{align}
\forall a \in A : (a,v) \in R \nonumber
\end{align}
\end{definition}

The satisfaction of an IKC $C$ by a set $A$ will be noted $C(A)$ in this paper. Consider an ordered set $U$, an IKC $C = (R,v)$ on $U$ and a set $A \subseteq U$. Due to the uncertainty inherent to $v$, it is uncertain whether $A$ satisfies $C$ or not. The degree of possibility $\Pos(C(A))$ that $A$ satisfies $C$ and the degree of necessity $\Nec(C(A))$ that $A$ satisfies $C$, can be calculated as follows~\cite{Pons2013ijufkbs}:

\begin{align}
\Pos(C(A)) & = \min_{a \in A}\left(\sup_{(a,w) \in R}\pi_{X_{v}}(w)\right) \label{ill-known-pos}\\
\Nec(C(A)) & = \min_{a \in A}\left(\inf_{(a,w) \notin R} 1-\pi_{X_{v}}(w)\right) \label{ill-known-nec}
\end{align}

\begin{table}[b]
\caption{The translations of Allen relationships to the IKC framework}
\centering
\begin{tabular}{|c|c|c|}
\hline
\textbf{Allen Relationship} & \textbf{Constraints} & \textbf{Combination} \\
\hline
\hline
I before J & $C_1\stackrel{\triangle}{=} \left(<,s\right)$ & $C_1(I)$ \\
\hline
\multirow{2}{*}
{I equal J} & $C_1\stackrel{\triangle}{=} \left(\geq,s\right)$, $C_2\stackrel{\triangle}{=} \left(\neq,s\right)$ & $C_1(I)\wedge$ $\neg C_2(I)\wedge$\\
& $C_3\stackrel{\triangle}{=} \left(\leq,e\right)$, $C_4\stackrel{\triangle}{=} \left(\neq,e\right)$ & $C_3(I)\wedge$ $\neg C_4(I)$\\
\hline
I meets J & $C_1\stackrel{\triangle}{=} \left(\leq,s\right)$ $C_2\stackrel{\triangle}{=} \left(\neq,s\right)$ & $C_1(I)\wedge$ $\neg C_2(I)$\\
\hline
I overlaps J & $C_1\stackrel{\triangle}{=} \left(<,e\right)$, $C_2\stackrel{\triangle}{=} \left(\leq,s\right)$, $C_3\stackrel{\triangle}{=} \left(\geq,s\right)$ & $C_1(I)\wedge$ $\neg C_2(I)\wedge$ $\neg C_3(I)$\\
\hline
\multirow{2}{*}
{I during J} & $C_1\stackrel{\triangle}{=} \left(>,s\right)$, $C_2\stackrel{\triangle}{=} \left(\leq,e\right)$ & $\big(C_1(I)\wedge$ $ C_2(I)\big)\vee$\\
 & $C_3\stackrel{\triangle}{=} \left(\geq,s\right)$, $C_4\stackrel{\triangle}{=} \left(<,e\right)$ & $\big(C_3(I)\wedge$ $C_4(I)\big)$\\
\hline
I starts J & $C_1\stackrel{\triangle}{=} \left(\geq,s\right)$, $C_2\stackrel{\triangle}{=} \left(\neq,s\right)$ & $C_1(I)\wedge$, $\neg C_2(I)$\\
\hline
I finishes J & $C_1\stackrel{\triangle}{=} \left(\leq,e\right)$, $C_2\stackrel{\triangle}{=} \left(\neq,e\right)$ & $C_1(I)\wedge$ $\neg C_2(I)$\\
\hline
\end{tabular}
\label{tab:allen-relations}
\end{table}

Given an ordered set $U$, degrees of possibility and necessity that a set $A \subseteq U$ satisfies a boolean combination of IKC on $U$ can be found by using the possibilistic extensions of boolean operators `and' ($\wedge$), `or' ($\vee$) and `not' ($\neg$), as described in~\cite{Billiet2012ipmu},~\cite{Pons2012ipmu},~\cite{Pons2013ijufkbs}.

@@

\subsection{Bipolar Querying}
As mentioned before, humans may express their query preferences using both positive and negative query conditions~\cite{DeTre2010ieeetfs},~\cite{Matthe2011ijis}. If the semantics of these conditions are non-symmetric, meaning that the positive preferences can not be derived from the negative or vice versa, the bipolarity in this query is called \emph{heterogenous}~\cite{Matthe2011ijis}. The presented work will concern only such heterogenous query bipolarity.

A query usually takes the form of a boolean combination of elementary query conditions. Every elementary query condition then expresses the user's demands concerning a single attribute. Bipolarity in a query can either be specified between or inside the elementary query conditions. In~\cite{Matthe2011ijis}, it is shown that combining both approaches makes no sense and that the approach where bipolarity is specified inside elementary query conditions, using intuitionistic fuzzy sets~\cite{Atanassov1986fss}, is a more intuitive one. In the presented work, only the latter approach is used. In this approach, elementary query conditions express both what is accepted and what is not accepted by the query, at once, and are called \emph{bipolar query conditions} (BQC)~\cite{Matthe2011ijis}.

Consider a relation attribute $A$. Let $dom_{A}$ be the domain of $A$'s data type, let $\mu_{c_{A}}$ and $\nu_{c_{A}}$ be membership functions from $dom_{A}$ to the unit interval $\left[0, 1\right]$, where $\mu_{c_{A}}(x)$ represents to what extent $x \in dom_{A}$ is satisfactory and $\nu_{c_{A}}(x)$ to what extent $x$ is unsatisfactory to a user, then a BQC $c_{A}$ expressing this user's preferences concerning $A$ can be modelled by an Intuitionistic Fuzzy Set (IFS)~\cite{Atanassov1986fss} as~\cite{DeTre2010ieeetfs},~\cite{Matthe2011ijis}:
\vspace{-5pt}
\begin{equation}
c_{A} = \{(v, \mu_{c_{A}}(v), \nu_{c_{A}}(v)) : v \in dom_{A}\}
\end{equation}

Note that to allow overspecification of the user's preferences, the IFS's consistency condition can be relaxed, which means that there may exist values $v \in dom_{A}$ for which $\mu_{c_{A}}(v) + \nu_{c_{A}}(v) > 1$~\cite{DeTre2010ieeetfs},~\cite{Matthe2011ijis}.

@@

In general, the evaluation of a BQC $c_{A}$ on relation attribute $A$ for a tuple $r$ will result in a BSD $(s_{c_A}^r, d_{c_A}^r)$, which is calculated as follows. Let $r[A]$ denote the value of tuple $r$ for attribute $A$, then~\cite{Matthe2011ijis}:
\begin{equation}\label{eq:bsdeval}
(s_{c_A}^r, d_{c_A}^r)=(\mu_{c_A}(r[A]), \nu_{c_A}(r[A]))
\end{equation}

@@

\subsection{Valid-Time Relations Subject to Uncertainty}
The presented proposal will concern VTR where the VTI are IKVTI. Generally, such a VTR $R$ can be seen as constructed in the following way. Let $R$ have $n$ non-temporal attributes $A_{i}, 1 \leq i \leq n, i \in \mathbb{N}$ and two valid-time attributes $VST$ and $VET$. Every tuple $T$ of $R$ represents an object or concept version or state which is valid during the time period indicated by the tuple's IKVTI $I_T$. This IKVTI $I_T$ is now defined by two IKV, which respectively describe the starting and ending instants of $I_T$ and are represented by the tuple's VST, respectively VET values. The interpretation is that the version or state corresponding to a tuple was, is or will be valid during a period of time, but exactly which period this is intended to be, is unknown. Confidence about exactly which period is intended, is modelled by the tuple's IKVTI~\cite{Pons2012ijcis},~\cite{Pons2012ipmu}.

\vspace{-10pt}
\begin{table}[ht]
\caption{The example relation used in this paper}
\centering
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{ID} & \textbf{Author} & \textbf{VST} & \textbf{VET} \\
\hline
\hline
1 & Alo\"{i}sius & $\left[4/4/1208, 6/4/1208, 16/4/1208\right]$ & $\left[10/12/1208, 1/1/1209, 26/1/1209\right]$ \\
\hline
2 & Theofilus & $\left[2/4/1209, 12/4/1209, 22/4/1209\right]$ & $\left[21/12/1209, 1/1/1210, 21/1/1210\right]$ \\
\hline
3 & Gerardus & $\left[14/1/1209, 15/1/1209, 16/1/1209\right]$ & $\left[21/12/1209, 15/1/1210, 25/1/1210\right]$ \\
\hline
4 & Euforius & $\left[21/12/1210, 1/1/1211, 11/1/1211\right]$ & $\left[21/12/1211, 1/1/1212, 11/1/1212\right]$ \\
\hline
5 & Ambrosius & $\left[11/12/1213, 21/12/1213, 15/1/1214\right]$ & $\left[9/10/1216, 10/10/1216, 15/10/1216\right]$ \\
\hline
6 & Alo\"{i}sius & $\left[21/12/1213, 1/1/1214, 11/1/1214\right]$ & $\left[9/6/1217, 9/6/1217, 12/6/1217\right]$ \\
\hline
7 & Gerardus & $\left[29/12/1214, 1/1/1215, 8/1/1215\right]$ & $\left[9/6/1217, 10/6/1217, 12/6/1217\right]$ \\
\hline
\end{tabular}
\label{tab:ex-relation}
\end{table}
\vspace{-15pt}

@@

\vspace{-10pt}
\begin{align}
\pi(x) =
\begin{cases}
\frac{x - d_1}{d_2 - d_1}, & \text{if } d_1 \leq x < d_2 \\
\frac{d_3 - x}{d_3 - d_2}, & \text{if } d_2 \leq x \leq d_3 \\
0, & \text{else}
\end{cases}
\end{align}
\vspace{-15pt}
